---
title: "Project 4"
---

```{r setup, include=FALSE}
library(tidyverse)
library(colorspace)
library(dplyr)
library(broom)
knitr::opts_chunk$set(echo = TRUE)
```

This is the dataset you will be working with:

```{r message = FALSE}
lemurs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-08-24/lemur_data.csv')

lemurs

length(unique(lemurs[["dlc_id"]]))
```

More information about the dataset can be found here:
https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-08-24 and https://www.nature.com/articles/sdata201419.


**Question:** *What are the differences between taxons when looking at their expected gestation length, litter size, age of conception of the mother and father, and weight?*

**Introduction:** *The Duke Lemur Center houses over 200 lemurs across 14 species – the most diverse population of lemurs on Earth, outside their native Madagascar. With information on 2270 different lemurs acquired by the DLC, this study will investigate differences between the taxons of lemurs with respect to features of their birth. We will investigate their expected gestation length (expected_gestation_d), litter size (litter_size), age of conception for the mother and father (sire_age_at_concep_y and dam_age_at_concep_y), and their weight (weight_g).*

**Approach:** *We only want the latest lemur information, so we'll keep only the rows which have the latest weight date per each DLC_id. Then we'll do PCA stuff: scale our numeric columns, perform PCA, calculate the components, and plot the PCs, rotation matrix, and variance explained.*

**Analysis:**

```{r fig.height=10, fig.width=10}
# getting the latest weight date rows per dlc id
lemurs_latest <- lemurs %>% 
  group_by(dlc_id) %>% 
  mutate(taxon = recode(taxon, CMED = "Fat-tailed dwarf lemur", DMAD = "Aye-aye", EALB = "White-fronted brown lemur", ECOL = "Collared brown lemur", ECOR = "Crowned lemur", EFLA = "Blue-eyed black lemur", EFUL = "Common brown lemur", EMAC = "Black lemur", EMON = "Mongoose lemur", ERUB = "Red-bellied lemur", ERUF = "Red-fronted brown lemur", ESAN = "Sanford’s brown lemur", EUL = "Hybrid Elemur", GMOH = "Mohol bushbaby", HGG = "Eastern lesser bamboo lemur", LCAT = "Ring-tailed lemur", LTAR = "Slender loris", MMUR = "Gray mouse lemur", MZAZ = "Northern giant mouse lemur", NCOU = "Slow loris", NPYG = "Pygmy slow loris", OGG = "Northern greater galago", PCOQ = "Coquerel’s sifaka", PPOT = "Potto", VAR = "Hybrid Varecia", VRUB = "Red ruffed lemur", VVV = "Black-and-white ruffed lemur")) %>% #more useful names from the github page
  slice(which.max(as.Date(weight_date, '%m/%d/%Y')))

# ungroup acting weird, had to make a temp variable. 
temp <- lemurs_latest %>% 
  ungroup() %>%
  select(dlc_id, taxon, expected_gestation, litter_size, sire_age_at_concep_y, dam_age_at_concep_y, weight_g) %>%
  na.omit #can't have missing values in PCA, only lost 680 lemurs out of 2270 or 30%.
  

pca_fit <- temp %>% 
  select(expected_gestation, litter_size, sire_age_at_concep_y, dam_age_at_concep_y, weight_g) %>%
  scale() %>%
  prcomp()
  

pca_fit %>%
  # add PCs to the original dataset
  augment(temp) %>%
  ggplot(aes(.fittedPC1, .fittedPC2)) +
  geom_point(size=1, aes(color=taxon)) + 
  facet_wrap(~taxon) +
  guides(colour=FALSE) 

pca_fit %>%
  # add PCs to the original dataset
  augment(temp) %>%
  ggplot(aes(.fittedPC2, .fittedPC3)) +
  geom_point(size=1, aes(color=taxon)) + 
  facet_wrap(~taxon) +
  guides(colour=FALSE) 

```
```{r}
arrow_style <- arrow(
  angle = 20, length = grid::unit(8, "pt"),
  ends = "first", type = "closed"
)
pca_fit %>%
  # extract rotation matrix
  tidy(matrix = "rotation") %>%
  pivot_wider(
    names_from = "PC", values_from = "value",
    names_prefix = "PC"
  ) %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(
    xend = 0, yend = 0,
    arrow = arrow_style
  ) +
  geom_text(aes(label = column), hjust = 1) +
  xlim(-1.5, 0.5) + ylim(-1, 1) + 
  coord_fixed()

pca_fit %>%
  # extract rotation matrix
  tidy(matrix = "rotation") %>%
  pivot_wider(
    names_from = "PC", values_from = "value",
    names_prefix = "PC"
  ) %>%
  ggplot(aes(PC2, PC3)) +
  geom_segment(
    xend = 0, yend = 0,
    arrow = arrow_style
  ) +
  geom_text(aes(label = column), hjust = 1) +
  xlim(-1.5, 0.5) + ylim(-1, 1) + 
  coord_fixed()


pca_fit %>%
  # extract eigenvalues
  tidy(matrix = "eigenvalues") %>%
  ggplot(aes(PC, percent)) + 
  geom_col() + 
  scale_x_continuous(
    # create one axis tick per PC
    breaks = 1:6
  ) +
  scale_y_continuous(
    name = "variance explained",
    # format y axis ticks as percent values
    label = scales::label_percent(accuracy = 1)
  )
```

**Discussion:** *Discussion of the plots in reverse order will be helpful. Let's start with the explained variance plot. Principal components 1, 2, and 3 cover around 80% of the explained variance in the dataset. So it's best to leave the dicussion to these two PCs. All variables except for litter size contribute negatively to PC1, showing that higher litter size is an indicator of a lower gestation period, weight, and age of parents. PC2 shows how litter size has an effect on gestation periods, indicating the same relationship found in PC1. PC3 shows another relationship between the age of parents and the weight. Typically, the older the parents, the leaner the lemur.*

*Let's move onto answering the question, given our interpretation of the principal components. When looking at PC1 plotted against PC2, you immediately notice "streaks" of observations that lie in the direction of the parents' ages and weight of the lemur. The groups (or "streaks") represent lemurs that have similarly sized litters (and therefore gestation periods), spread along their weights and ages of their parents.	Coquerel’s sifaka lemurs tend to stick to one litter size, while Black-and-white ruffed lemurs have varying litter sizes.  Weight will pull the points towards the third quadrant, so Northern giant mouse lemurs tend to be heavier than the Red ruffed lemur, for example. (It's slightly humorous that the "giant mouse" lemur is actually quite small compared to the average lemur.*

*Moving onto PC2 vs PC3, agian we see the "streaking" litter sizes effect, but we can observe some differences in parental age (age of fertility) for different lemurs. Older parents will pull the points towards the second quadrant, so fat tailed dwarf lemurs tend to have babies at an older age compared to the rest. We can also see which lemurs are fat by looking at points pulled towards the third quadrant, Coquerel’s sifaka tend to be heavier compared to the other lemurs.*


